---
title: "MTG Card Explorer"
author: "Eli Crowder & Evan Donch"
format:
  dashboard:
    theme: darkly
    nav-buttons: []
    css: dashboard.css
---

```{r setup, include=FALSE}
# ── Packages ──────────────────────────────────────────────────────────────────
# tidyverse : data wrangling (dplyr, tidyr) and string tools (stringr)
# jsonlite  : parse the JSON response from the Scryfall API
# plotly    : interactive charts (hover, zoom, pan)
# DT        : interactive sortable/filterable HTML tables
# crosstalk : links filter widgets to a DT table without needing Shiny
library(tidyverse)
library(jsonlite)
library(plotly)
library(DT)
library(crosstalk)


# ── Step 1: Pull the Scryfall bulk data URL ───────────────────────────────────
# Scryfall exposes a /bulk-data endpoint that lists available download files.
# We first fetch that index, then grab the download URL for "default_cards",
# which contains one entry per card printing (non-foil, English).
bulk_info    <- fromJSON("https://api.scryfall.com/bulk-data")
download_url <- bulk_info$data$download_uri[
  bulk_info$data$type == "default_cards"
]


# ── Step 2: Download and trim the full card list ──────────────────────────────
# The bulk JSON is large (~100 MB) but flatten = TRUE turns nested lists into
# dot-separated column names, making it easy to work with as a data frame.
# We immediately select only the columns we actually need to keep memory low.
mtg_small <- fromJSON(download_url, flatten = TRUE) |>
  as_tibble() |>
  select(
    oracle_id,              # Unique ID for the card concept (same across printings)
    name,                   # Card name
    mana_cost,              # Mana cost string, e.g. "{2}{U}{B}"
    cmc,                    # Converted Mana Cost — total mana as a number
    type_line,              # Full type string, e.g. "Legendary Creature — Dragon"
    colors,                 # List-column of color codes: W, U, B, R, G
    color_identity,         # Color identity (includes activated ability costs)
    set,                    # 3-letter set code, e.g. "lea" for Alpha
    rarity,                 # common / uncommon / rare / mythic / special / bonus
    prices.usd,             # Lowest USD price across all non-foil market listings
    legalities.standard,    # "legal", "not_legal", "banned", or "restricted"
    legalities.modern,
    legalities.commander
  )


# ── Step 3: Clean and deduplicate ────────────────────────────────────────────
# Many cards appear in multiple sets (reprints). We want one row per unique
# card concept, so we group by oracle_id and keep the cheapest printing.
# This avoids double-counting the same card under different set codes.
mtg_cheapest <- mtg_small |>

  # Convert prices from character to numeric (NAs introduced for missing prices)
  mutate(price = as.numeric(prices.usd)) |>

  # Drop cards with no USD price data — these are typically digital-only or
  # promo cards that don't have a secondary market listing
  filter(!is.na(price)) |>

  # Keep only the cheapest printing per oracle_id
  group_by(oracle_id) |>
  slice_min(price, with_ties = FALSE) |>
  ungroup() |>

  mutate(
    # Collapse the list-column of color codes into one sorted string.
    # sort() ensures consistent ordering (e.g., always "UB" not "BU").
    # Cards with no colors (lands, artifacts) get labeled "Colorless".
    color_str = sapply(colors, function(x) {
      if (length(x) == 0) "Colorless" else paste(sort(x), collapse = "")
    }),

    # Count how many colors a card has — used to detect multicolor cards
    color_count = sapply(colors, length),

    # Broad single-label color category used in overview charts.
    # Any card with 2+ colors is "Multicolor" regardless of which colors.
    color_label = case_when(
      color_count == 0  ~ "Colorless",
      color_count >= 2  ~ "Multicolor",
      color_str == "W"  ~ "White",
      color_str == "U"  ~ "Blue",
      color_str == "B"  ~ "Black",
      color_str == "R"  ~ "Red",
      color_str == "G"  ~ "Green",
      TRUE              ~ "Other"
    ),

    # Assign one primary card type per card.
    # str_detect() is used because type_line can contain multiple types,
    # e.g. "Artifact Creature" — Creature takes precedence here.
    card_type = case_when(
      str_detect(type_line, "Creature")     ~ "Creature",
      str_detect(type_line, "Instant")      ~ "Instant",
      str_detect(type_line, "Sorcery")      ~ "Sorcery",
      str_detect(type_line, "Enchantment")  ~ "Enchantment",
      str_detect(type_line, "Artifact")     ~ "Artifact",
      str_detect(type_line, "Land")         ~ "Land",
      str_detect(type_line, "Planeswalker") ~ "Planeswalker",
      TRUE                                  ~ "Other"
    ),

    # Re-encode rarity as an ordered factor so charts display in the correct
    # gameplay order (common → uncommon → rare → mythic) not alphabetically
    rarity = factor(
      rarity,
      levels = c("common", "uncommon", "rare", "mythic", "special", "bonus")
    ),

    # Full human-readable color combination used in the Card Browser filter.
    # Unlike color_label, this distinguishes "Blue/Black" from plain "Blue",
    # so the filter dropdown shows exact multi-color combinations.
    color_combo = sapply(colors, function(x) {
      if (length(x) == 0) return("Colorless")
      lookup <- c(W = "White", U = "Blue", B = "Black", R = "Red", G = "Green")
      paste(lookup[sort(x)], collapse = "/")
    })
  )


# ── Color palettes ────────────────────────────────────────────────────────────
# Named vectors so we can look up colors by name in plot_ly calls.
# Colors are chosen to match the official MTG card frame colors.

# One color per MTG identity
mtg_pal <- c(
  White     = "#f0e6c0",   # Cream/parchment — the Plains color
  Blue      = "#1e6eb5",   # Island blue
  Black     = "#8a7a9b",   # Swamp purple-gray (pure black is hard to see)
  Red       = "#d44239",   # Mountain red
  Green     = "#2a7a45",   # Forest green
  Colorless = "#a0a0a0",   # Neutral gray for artifacts/lands
  Multicolor = "#c8a032"   # Gold — the traditional multicolor frame color
)

# Rarity colors match the card border treatment in the actual game
rarity_pal <- c(
  common   = "#9e9e9e",   # Black border / gray
  uncommon = "#5dade2",   # Silver
  rare     = "#d4ac0d",   # Gold
  mythic   = "#e67e22",   # Orange-gold
  special  = "#8e44ad"    # Purple (used for Timeshifted etc.)
)

# Card type colors — loosely chosen to match thematic associations
type_pal <- c(
  Creature     = "#2a7a45",   # Green — living things
  Instant      = "#1e6eb5",   # Blue — reactive/tricky
  Sorcery      = "#d44239",   # Red — powerful/direct
  Enchantment  = "#8e44ad",   # Purple — magical/persistent
  Artifact      = "#95a5a6",  # Silver-gray — mechanical
  Land          = "#a0522d",  # Brown — terrain
  Planeswalker  = "#c8a032",  # Gold — iconic characters
  Other         = "#555555"   # Dark gray — catch-all
)


# ── Pre-computed summary values ───────────────────────────────────────────────
# Calculated once here so they can be referenced across all pages and inside
# inline R expressions in the markdown text cards (e.g. `r avg_price`).

total_cards <- nrow(mtg_cheapest)
avg_price   <- round(mean(mtg_cheapest$price), 2)
med_price   <- round(median(mtg_cheapest$price), 2)
most_exp    <- mtg_cheapest |> slice_max(price, n = 1, with_ties = FALSE)
total_sets  <- n_distinct(mtg_cheapest$set)

# Price tier counts — used in value boxes and inline text on Price Analysis page
n_cheap <- sum(mtg_cheapest$price < 1)    # Cards under $1
n_exp   <- sum(mtg_cheapest$price >= 10)  # Cards $10 or more
n_rare  <- sum(mtg_cheapest$price >= 50)  # Cards $50 or more
```

# About This Dataset

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Unique Cards"
# total_cards is the number of rows after deduplication by oracle_id
list(icon = "collection", color = "primary",
     value = format(total_cards, big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Avg Price (USD)"
# Mean price across all unique cards (cheapest printing each)
list(icon = "currency-dollar", color = "warning",
     value = paste0("$", avg_price))
```

```{r}
#| content: valuebox
#| title: "Most Expensive"
# Card with the single highest market price in the dataset
list(icon = "trophy", color = "danger",
     value = paste0("$", round(most_exp$price, 2)),
     subtitle = most_exp$name)
```

```{r}
#| content: valuebox
#| title: "Sets Represented"
# Count of distinct 3-letter set codes in the deduplicated dataset
list(icon = "layers", color = "success",
     value = format(total_sets, big.mark = ","))
```

## Row {height="80%"}

### Column {width="40%"}

::: {.card}
### About This Dashboard

This dashboard explores **Magic: The Gathering** card data sourced from the
[Scryfall](https://scryfall.com) bulk data API — a comprehensive, regularly
updated database of every MTG card ever printed.

**What is in the dataset?**

Each row represents a unique card (identified by its `oracle_id`), keeping
only the cheapest available printing when multiple exist. The key variables we
analyze are:

| Variable | Description |
|---|---|
| `name` | Card name |
| `cmc` | Converted Mana Cost — total mana to cast |
| `colors` | Color identity (W, U, B, R, G) |
| `type_line` | Card type (Creature, Instant, etc.) |
| `rarity` | Common, Uncommon, Rare, Mythic, Special |
| `price` | Lowest USD market price across all printings |
| `legalities` | Whether the card is legal in Standard, Modern, or Commander |

**What does this dashboard explore?**

- **Price Analysis** — How are card prices distributed? What drives value?
- **Mana & Color** — How complex are cards by mana cost and color?
- **Format Legality** — What share of cards are legal in each format?
- **Card Browser** — Search and filter the full card database yourself.
:::

### Column {width="60%"}

```{r}
#| title: "Card Count by Color"

# Count cards per broad color category, then set factor order for the chart
color_counts <- mtg_cheapest |>
  count(color_label) |>
  mutate(color_label = factor(
    color_label,
    levels = c("White", "Blue", "Black", "Red", "Green", "Colorless", "Multicolor")
  ))

# Use mtg_pal to color each bar by the card's actual MTG color.
# cliponaxis = FALSE lets count labels render above bars without being clipped.
plot_ly(color_counts,
  x     = ~reorder(color_label, -n),
  y     = ~n,
  type  = "bar",
  marker = list(
    color = mtg_pal[as.character(color_counts$color_label)],
    line  = list(color = "#1a1a2e", width = 1)
  ),
  text          = ~format(n, big.mark = ","),
  textposition  = "outside",
  cliponaxis    = FALSE,
  hovertemplate = "<b>%{x}</b><br>Cards: %{y:,}<extra></extra>"
) |>
  layout(
    xaxis  = list(title = "Color",      color = "#c9d1d9", automargin = TRUE),
    yaxis  = list(title = "Card Count", color = "#c9d1d9", automargin = TRUE),
    paper_bgcolor = "#1a1a2e",
    plot_bgcolor  = "#1a1a2e",
    font   = list(color = "#c9d1d9"),
    margin = list(t = 40, b = 60, l = 60, r = 40)
  )
```

```{r}
#| title: "Card Count by Type"

# Count cards per type and sort ascending so the longest bar is at the top
# when we flip to a horizontal bar chart
type_counts <- mtg_cheapest |>
  count(card_type) |>
  arrange(n)

# Horizontal bar chart — easier to read long category names on the y-axis
plot_ly(type_counts,
  x           = ~n,
  y           = ~reorder(card_type, n),
  type        = "bar",
  orientation = "h",
  marker = list(
    color = type_pal[as.character(type_counts$card_type)],
    line  = list(color = "#1a1a2e", width = 1)
  ),
  text          = ~format(n, big.mark = ","),
  textposition  = "outside",
  cliponaxis    = FALSE,
  hovertemplate = "<b>%{y}</b><br>Cards: %{x:,}<extra></extra>"
) |>
  layout(
    xaxis  = list(title = "Card Count", color = "#c9d1d9", automargin = TRUE),
    yaxis  = list(title = "",           color = "#c9d1d9", automargin = TRUE),
    paper_bgcolor = "#1a1a2e",
    plot_bgcolor  = "#1a1a2e",
    font   = list(color = "#c9d1d9"),
    margin = list(t = 40, b = 60, l = 120, r = 60)
  )
```

# Price Analysis

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Median Price"
# Median is more meaningful than mean here because of extreme outliers
list(icon = "align-center", color = "info",
     value = paste0("$", med_price))
```

```{r}
#| content: valuebox
#| title: "Cards Under $1"
list(icon = "tag", color = "success",
     value    = format(n_cheap, big.mark = ","),
     subtitle = paste0(round(n_cheap / total_cards * 100), "% of all cards"))
```

```{r}
#| content: valuebox
#| title: "Cards Over $10"
list(icon = "gem", color = "warning",
     value    = format(n_exp, big.mark = ","),
     subtitle = paste0(round(n_exp / total_cards * 100), "% of all cards"))
```

```{r}
#| content: valuebox
#| title: "Cards Over $50"
list(icon = "trophy-fill", color = "danger",
     value    = format(n_rare, big.mark = ","),
     subtitle = paste0(round(n_rare / total_cards * 100), "% of all cards"))
```

## Row {height="40%"}

### Column {width="65%"}

```{r}
#| title: "Price Distribution (USD) — log scale shows the full spread"

# A linear histogram is not useful here because ~80% of cards cost under $1,
# which creates one massive bar at zero and nothing visible elsewhere.
# Plotting on a log10 scale spreads the distribution out so we can see
# how many cards fall into each price tier: cents, dollars, tens, hundreds.
#
# We log10-transform the price before plotting, then relabel the x-axis
# tick marks as actual dollar amounts so the chart stays readable.

price_log <- mtg_cheapest |>
  filter(price > 0) |>                      # log(0) is undefined — drop free cards
  mutate(log_price = log10(price))

# Custom tick positions and their dollar-amount labels
tick_vals   <- c(-2, -1, 0, 1, 2, 3)        # log10 values: 0.01, 0.10, 1, 10, 100, 1000
tick_labels <- c("$0.01", "$0.10", "$1", "$10", "$100", "$1,000")

plot_ly(price_log,
  x    = ~log_price,
  type = "histogram",
  nbinsx = 80,
  marker = list(
    color = "#c8a032",
    line  = list(color = "#1a1a2e", width = 0.5)
  ),
  # Show the original dollar price in the tooltip, not the log value
  hovertemplate = "Price range: 10^%{x:.2f}<br>Count: %{y:,}<extra></extra>"
) |>
  layout(
    xaxis = list(
      title      = "Price (USD) — log scale",
      color      = "#c9d1d9",
      tickvals   = tick_vals,
      ticktext   = tick_labels,
      tickmode   = "array"
    ),
    yaxis = list(title = "Number of Cards", color = "#c9d1d9"),
    paper_bgcolor = "#1a1a2e",
    plot_bgcolor  = "#1a1a2e",
    font = list(color = "#c9d1d9")
  )
```

### Column {width="35%"}

::: {.card}
### Key Price Findings

The chart uses a **log scale** on the x-axis — each step to the right is 10×
more expensive. This reveals the true shape of the distribution, which is
impossible to see on a linear scale because so many cards cost under \$1.

The bell-shaped curve on a log scale tells us that card prices follow a
**log-normal distribution** — most cards cluster around a central value, with
tails of very cheap and very expensive cards on either side.

**Summary statistics:**

- **Mean price:** \$`r avg_price` — pulled up by expensive outliers
- **Median price:** \$`r med_price` — a better measure of the "typical" card
- **`r round(n_cheap / total_cards * 100)`%** of cards cost under \$1
- Only **`r round(n_rare / total_cards * 100)`%** cost \$50 or more

The large gap between mean and median confirms that a small number of
extremely valuable cards skew the average significantly upward.
:::

## Row {height="40%"}

### Column

```{r}
#| title: "Average Price by Rarity"

# Remove "bonus" rarity — these are special promo cards that don't fit
# the standard rarity hierarchy and would distort the comparison
rarity_avg <- mtg_cheapest |>
  filter(!rarity %in% c("bonus")) |>
  group_by(rarity) |>
  summarize(avg_price = round(mean(price, na.rm = TRUE), 2), .groups = "drop") |>
  # Re-apply factor ordering so bars appear in gameplay order, not alphabetically
  mutate(rarity = factor(
    rarity,
    levels = c("common", "uncommon", "rare", "mythic", "special")
  ))

plot_ly(rarity_avg,
  x             = ~rarity,
  y             = ~avg_price,
  type          = "bar",
  marker        = list(
    color = rarity_pal[as.character(rarity_avg$rarity)],
    line  = list(color = "#1a1a2e", width = 1)
  ),
  text          = ~paste0("$", avg_price),
  textposition  = "outside",
  cliponaxis    = FALSE,
  hovertemplate = "<b>%{x}</b><br>Avg Price: $%{y:.2f}<extra></extra>"
) |>
  layout(
    xaxis  = list(title = "Rarity",         color = "#c9d1d9", automargin = TRUE),
    yaxis  = list(title = "Avg Price (USD)", color = "#c9d1d9", automargin = TRUE),
    paper_bgcolor = "#1a1a2e",
    plot_bgcolor  = "#1a1a2e",
    font   = list(color = "#c9d1d9"),
    margin = list(t = 40, b = 60, l = 60, r = 40)
  )
```

### Column

```{r}
#| title: "Average Price by Card Type"

# arrange() by avg_price before plotting so the horizontal bar chart
# naturally sorts from cheapest (bottom) to most expensive (top)
type_avg <- mtg_cheapest |>
  group_by(card_type) |>
  summarize(avg_price = round(mean(price, na.rm = TRUE), 2), .groups = "drop") |>
  arrange(avg_price)

plot_ly(type_avg,
  x           = ~avg_price,
  y           = ~reorder(card_type, avg_price),
  type        = "bar",
  orientation = "h",
  marker      = list(
    color = type_pal[as.character(type_avg$card_type)],
    line  = list(color = "#1a1a2e", width = 1)
  ),
  text          = ~paste0("$", avg_price),
  textposition  = "outside",
  cliponaxis    = FALSE,
  hovertemplate = "<b>%{y}</b><br>Avg Price: $%{x:.2f}<extra></extra>"
) |>
  layout(
    xaxis  = list(title = "Avg Price (USD)", color = "#c9d1d9", automargin = TRUE),
    yaxis  = list(title = "",                color = "#c9d1d9", automargin = TRUE),
    paper_bgcolor = "#1a1a2e",
    plot_bgcolor  = "#1a1a2e",
    font   = list(color = "#c9d1d9"),
    margin = list(t = 40, b = 60, l = 120, r = 60)
  )
```

# Mana & Color

## Row {height="15%"}

::: {.card}
### What is Converted Mana Cost (CMC)?

Every MTG card requires **mana** to cast — the game's primary resource.
**Converted Mana Cost (CMC)** is the total amount of mana required, regardless
of color. A CMC of 1 is a cheap, fast card; a CMC of 7+ is a powerful late-game
spell. Understanding CMC helps reveal the design philosophy behind the game:
aggressive strategies rely on low-CMC cards, while control decks can afford
higher costs. The scatter plot below lets you explore whether higher mana cost
actually translates to higher market value.
:::

## Row {height="42%"}

### Column {width="60%"}

```{r}
#| title: "CMC Distribution — How Complex Are Cards?"

# Filter to CMC <= 16 to exclude a handful of extreme outlier cards
# (e.g., some Un-set joke cards have absurdly high CMC values)
cmc_counts <- mtg_cheapest |>
  count(cmc) |>
  filter(cmc <= 16)

# dtick = 1 forces the x-axis to show every integer CMC value
plot_ly(cmc_counts,
  x             = ~cmc,
  y             = ~n,
  type          = "bar",
  marker        = list(color = "#1e6eb5",
                       line  = list(color = "#1a1a2e", width = 0.5)),
  hovertemplate = "CMC %{x}<br>Cards: %{y:,}<extra></extra>"
) |>
  layout(
    xaxis  = list(title = "Converted Mana Cost (CMC)", color = "#c9d1d9", dtick = 1),
    yaxis  = list(title = "Number of Cards",            color = "#c9d1d9"),
    paper_bgcolor = "#1a1a2e",
    plot_bgcolor  = "#1a1a2e",
    font   = list(color = "#c9d1d9")
  )
```

### Column {width="40%"}

```{r}
#| title: "Average CMC by Card Type"

# Lands have CMC 0 (they don't cost mana to play), which will anchor
# the bottom of the chart — a useful reference point
cmc_by_type <- mtg_cheapest |>
  group_by(card_type) |>
  summarize(avg_cmc = round(mean(cmc, na.rm = TRUE), 2), .groups = "drop") |>
  arrange(avg_cmc)

plot_ly(cmc_by_type,
  x           = ~avg_cmc,
  y           = ~reorder(card_type, avg_cmc),
  type        = "bar",
  orientation = "h",
  marker      = list(
    color = type_pal[as.character(cmc_by_type$card_type)],
    line  = list(color = "#1a1a2e", width = 1)
  ),
  text          = ~avg_cmc,
  textposition  = "outside",
  cliponaxis    = FALSE,
  hovertemplate = "<b>%{y}</b><br>Avg CMC: %{x:.2f}<extra></extra>"
) |>
  layout(
    xaxis  = list(title = "Avg CMC", color = "#c9d1d9", automargin = TRUE),
    yaxis  = list(title = "",        color = "#c9d1d9", automargin = TRUE),
    paper_bgcolor = "#1a1a2e",
    plot_bgcolor  = "#1a1a2e",
    font   = list(color = "#c9d1d9"),
    margin = list(t = 40, b = 60, l = 120, r = 60)
  )
```

## Row {height="43%"}

```{r}
#| title: "Price vs. CMC — Hover over points for card details (sample of 5,000 cards, prices ≤ $100)"

# Plotting all 30 000+ cards as scatter points would freeze the browser.
# We take a random sample of 5 000 cards and cap price at $100 to focus
# on the bulk of the data. set.seed() makes the sample reproducible.
set.seed(42)
scatter_data <- mtg_cheapest |>
  filter(price <= 100) |>
  slice_sample(n = min(5000, nrow(mtg_cheapest))) |>
  # Re-apply factor so the legend displays colors in a consistent order
  mutate(color_label = factor(
    color_label,
    levels = c("White", "Blue", "Black", "Red", "Green", "Colorless", "Multicolor")
  ))

# Color each point by the card's MTG color using mtg_pal.
# hoverinfo = "text" suppresses plotly's default tooltip so only our
# custom HTML string is shown when the user hovers.
plot_ly(scatter_data,
  x          = ~cmc,
  y          = ~price,
  color      = ~color_label,
  colors     = mtg_pal,
  type       = "scatter",
  mode       = "markers",
  marker     = list(opacity = 0.5, size = 5),
  text       = ~paste0(
    "<b>", name, "</b><br>",
    "CMC: ",    cmc, "<br>",
    "Price: $", round(price, 2), "<br>",
    "Type: ",   card_type, "<br>",
    "Rarity: ", rarity
  ),
  hoverinfo  = "text"
) |>
  layout(
    xaxis  = list(title = "Converted Mana Cost", color = "#c9d1d9"),
    yaxis  = list(title = "Price (USD)",          color = "#c9d1d9"),
    legend = list(font  = list(color = "#c9d1d9")),
    paper_bgcolor = "#1a1a2e",
    plot_bgcolor  = "#1a1a2e",
    font   = list(color = "#c9d1d9")
  )
```

# Format Legality

## Row {height="15%"}

::: {.card}
### MTG's Major Competitive Formats

MTG has several **formats** that determine which cards are legal to play competitively.
**Standard** rotates yearly, only allowing cards from recent sets — keeping the
meta fresh but making cards obsolete. **Modern** allows cards from 2003 onward,
creating a much larger card pool. **Commander** is the most popular casual format,
where nearly every non-banned card is legal and players build 100-card decks around
a legendary creature. Because Standard bans more cards, legal cards there tend to
be more recently printed and often more expensive due to active demand.
:::

## Row {height="42%"}

### Column

```{r}
#| title: "Standard Legality"

# Count how many cards fall into each legality status for Standard.
# The four possible values are: legal, not_legal, banned, restricted.
std <- mtg_cheapest |>
  count(legalities.standard) |>
  rename(status = legalities.standard)

plot_ly(std,
  labels        = ~status,
  values        = ~n,
  type          = "pie",
  marker        = list(
    colors = c("#2a7a45", "#d44239", "#8b949e", "#c8a032"),
    line   = list(color = "#1a1a2e", width = 2)
  ),
  textinfo      = "label+percent",
  hovertemplate = "<b>%{label}</b><br>Cards: %{value:,}<br>%{percent}<extra></extra>"
) |>
  layout(
    paper_bgcolor = "#1a1a2e",
    font   = list(color = "#c9d1d9"),
    legend = list(font = list(color = "#c9d1d9"))
  )
```

### Column

```{r}
#| title: "Modern Legality"
mod <- mtg_cheapest |>
  count(legalities.modern) |>
  rename(status = legalities.modern)

plot_ly(mod,
  labels        = ~status,
  values        = ~n,
  type          = "pie",
  marker        = list(
    colors = c("#2a7a45", "#d44239", "#8b949e", "#c8a032"),
    line   = list(color = "#1a1a2e", width = 2)
  ),
  textinfo      = "label+percent",
  hovertemplate = "<b>%{label}</b><br>Cards: %{value:,}<br>%{percent}<extra></extra>"
) |>
  layout(
    paper_bgcolor = "#1a1a2e",
    font   = list(color = "#c9d1d9"),
    legend = list(font = list(color = "#c9d1d9"))
  )
```

### Column

```{r}
#| title: "Commander Legality"
cmd <- mtg_cheapest |>
  count(legalities.commander) |>
  rename(status = legalities.commander)

plot_ly(cmd,
  labels        = ~status,
  values        = ~n,
  type          = "pie",
  marker        = list(
    colors = c("#2a7a45", "#d44239", "#8b949e", "#c8a032"),
    line   = list(color = "#1a1a2e", width = 2)
  ),
  textinfo      = "label+percent",
  hovertemplate = "<b>%{label}</b><br>Cards: %{value:,}<br>%{percent}<extra></extra>"
) |>
  layout(
    paper_bgcolor = "#1a1a2e",
    font   = list(color = "#c9d1d9"),
    legend = list(font = list(color = "#c9d1d9"))
  )
```

## Row {height="43%"}

### Column {width="50%"}

```{r}
#| title: "Average Price of Legal Cards by Format"

# Stack all three formats into one long data frame so we can group_by format.
# We only keep rows where status == "legal" to compare prices of
# playable cards rather than banned/not-legal ones.
legality_price <- bind_rows(
  mtg_cheapest |> mutate(format = "Standard",  status = legalities.standard),
  mtg_cheapest |> mutate(format = "Modern",    status = legalities.modern),
  mtg_cheapest |> mutate(format = "Commander", status = legalities.commander)
) |>
  filter(status == "legal") |>
  group_by(format) |>
  summarize(
    avg_price = round(mean(price, na.rm = TRUE), 2),
    n_cards   = n(),
    .groups   = "drop"
  )

plot_ly(legality_price,
  x             = ~format,
  y             = ~avg_price,
  type          = "bar",
  marker        = list(
    color = c("#2a7a45", "#1e6eb5", "#c8a032"),
    line  = list(color = "#1a1a2e", width = 1)
  ),
  # Show both the avg price and card count in the bar label
  text          = ~paste0("$", avg_price, " (", format(n_cards, big.mark = ","), " cards)"),
  textposition  = "outside",
  cliponaxis    = FALSE,
  hovertemplate = "<b>%{x}</b><br>Avg Price: $%{y:.2f}<extra></extra>"
) |>
  layout(
    xaxis  = list(title = "Format",          color = "#c9d1d9", automargin = TRUE),
    yaxis  = list(title = "Avg Price (USD)", color = "#c9d1d9", automargin = TRUE),
    paper_bgcolor = "#1a1a2e",
    plot_bgcolor  = "#1a1a2e",
    font   = list(color = "#c9d1d9"),
    margin = list(t = 60, b = 60, l = 60, r = 40)
  )
```

### Column {width="50%"}

::: {.card}
### What the Legality Data Shows

```{r}
# Compute legality percentages inline so they can be referenced in the
# markdown text below using `r std_legal_pct` etc.
std_legal_pct <- round(
  sum(mtg_cheapest$legalities.standard  == "legal") / total_cards * 100
)
mod_legal_pct <- round(
  sum(mtg_cheapest$legalities.modern    == "legal") / total_cards * 100
)
cmd_legal_pct <- round(
  sum(mtg_cheapest$legalities.commander == "legal") / total_cards * 100
)
```

- Only **`r std_legal_pct`%** of all cards are Standard-legal, reflecting
  how aggressively that format rotates old sets out.
- **`r mod_legal_pct`%** are Modern-legal — a much larger pool drawing from
  over 20 years of card printings.
- **`r cmd_legal_pct`%** are Commander-legal, making it the most inclusive
  format by far.

Standard-legal cards command the **highest average prices** because they are
in active competitive demand. Once a card rotates out of Standard, its price
typically drops unless it sees play in Modern or Commander.

This shows that **format legality is one of the strongest drivers of card
value** — not just rarity alone.
:::

# Summary & Resources

## Row {height="100%"}

### Column {width="60%"}

::: {.card}
### Key Findings

This dashboard analyzed **`r format(total_cards, big.mark=",")`** unique Magic:
The Gathering cards sourced from the Scryfall database. Here is a summary of
what we found:

**Price**

- Card prices are extremely right-skewed. The median price is only
  \$`r med_price`, but the mean is \$`r avg_price` due to a small number of
  very expensive cards pulling the average up.
- `r round(n_cheap / total_cards * 100)`% of all cards cost under \$1, while
  just `r round(n_rare / total_cards * 100)`% cost \$50 or more.
- **Mythic rarity** cards have the highest average price, followed by Rare.
  Common cards are nearly worthless on the secondary market.
- **Planeswalkers** have the highest average price by card type, likely because
  they are almost always printed at Mythic rarity.

**Mana Cost & Complexity**

- Most cards have a CMC between 1 and 5, with CMC 2–3 being the most common
  range. This reflects the game's design principle of rewarding efficient,
  lower-cost cards.
- Lands have a CMC of 0 (they do not require mana to play), while Sorceries
  and Creatures tend to have higher average CMC than Instants.
- There is **no strong linear relationship** between CMC and price — a
  high-mana-cost card is not necessarily more valuable.

**Format Legality**

- Standard is the most restrictive format, with only
  `r round(sum(mtg_cheapest$legalities.standard == "legal") / total_cards * 100)`%
  of cards legal.
- Standard-legal cards carry the highest average price, confirming that
  **competitive demand drives value more than rarity alone**.
- Commander is the most inclusive format and the most popular casual format,
  accepting `r round(sum(mtg_cheapest$legalities.commander == "legal") / total_cards * 100)`%
  of all cards.
:::

### Column {width="40%"}

::: {.card}
### Further Reading & Resources

**Data Source**

- [Scryfall](https://scryfall.com) — The database powering this dashboard.
  Search any card, browse sets, and explore advanced filters.
- [Scryfall API Documentation](https://scryfall.com/docs/api) — Full
  reference for the bulk data API used to load this dataset.

**Learning More About MTG**

- [MTG Wiki — Card Types](https://mtg.fandom.com/wiki/Card_type) — Detailed
  explanation of every card type and subtype.
- [MTG Wiki — Converted Mana Cost](https://mtg.fandom.com/wiki/Mana_value) —
  How CMC (now called "Mana Value") is calculated.
- [MTGGoldfish](https://www.mtggoldfish.com) — In-depth price tracking,
  format meta analysis, and deck building tools.

**R & Data Visualization**

- [Quarto Dashboard Documentation](https://quarto.org/docs/dashboards/) —
  How this dashboard was built.
- [plotly for R](https://plotly.com/r/) — Reference for all interactive chart
  types used on this dashboard.
- [DT Package](https://rstudio.github.io/DT/) — The interactive table library
  used in the Card Browser.
- [crosstalk Package](https://rstudio.github.io/crosstalk/) — How the Card
  Browser filters are linked to the table without Shiny.
:::

# Card Browser

## Row

```{r}
#| title: "Browse Cards — Filter by Color, Type, Rarity, or Format (top 2,000 by price)"

# Select and rename only the columns useful to a browser/viewer.
# We cap at 2 000 rows (sorted by price descending) to keep the page
# responsive — rendering all 30 000+ rows at once in the browser is too slow.
browser_data <- mtg_cheapest |>
  select(
    Name        = name,
    `Mana Cost` = mana_cost,
    CMC         = cmc,
    Type        = card_type,
    Color       = color_combo,   # Exact combo e.g. "Blue/Black" not just "Multicolor"
    Rarity      = rarity,
    Set         = set,
    `Price USD` = price,
    Standard    = legalities.standard,
    Modern      = legalities.modern,
    Commander   = legalities.commander
  ) |>
  arrange(desc(`Price USD`)) |>
  slice_head(n = 2000)

# SharedData links the filter widgets and the DT table client-side.
# When the user picks a value in a filter_select, crosstalk tells the
# table to hide rows that don't match — no server or Shiny required.
shared <- SharedData$new(browser_data)

# Render all six dropdowns as a single HTML output inside one flex container.
# They MUST be in the same code chunk as the datatable() call so that
# crosstalk can register them all against the same SharedData object.
htmltools::tags$div(
  class = "filter-row",
  htmltools::tags$div(class = "filter-item",
    filter_select("ct_color",     "Color",     shared, ~Color,     multiple = TRUE)),
  htmltools::tags$div(class = "filter-item",
    filter_select("ct_type",      "Type",      shared, ~Type,      multiple = TRUE)),
  htmltools::tags$div(class = "filter-item",
    filter_select("ct_rarity",    "Rarity",    shared, ~Rarity,    multiple = TRUE)),
  htmltools::tags$div(class = "filter-item",
    filter_select("ct_standard",  "Standard",  shared, ~Standard,  multiple = TRUE)),
  htmltools::tags$div(class = "filter-item",
    filter_select("ct_modern",    "Modern",    shared, ~Modern,    multiple = TRUE)),
  htmltools::tags$div(class = "filter-item",
    filter_select("ct_commander", "Commander", shared, ~Commander, multiple = TRUE))
)

# Render the table, also bound to shared so it reacts to the filters above.
# formatCurrency() adds a "$" prefix and two decimal places to the price column.
datatable(
  shared,
  rownames = FALSE,
  class    = "stripe hover compact",
  options  = list(
    pageLength = 20,
    scrollX    = TRUE,
    dom        = "frtip"   # f=search, r=processing, t=table, i=info, p=pagination
  )
) |>
  formatCurrency("Price USD", currency = "$", digits = 2)
```
