---
title: "Exam 2"
author: "Eli Crowder"
format: pdf
---

# Results

## Table

The table below shows the average temperature, dissolved oxygen, and pH for each river. I first read in the rivers dataset and cleaned the variables so that the date was stored as a date, the site was treated as a river name, and the measurement variables were numeric. Then I grouped the data by river and used `summarize()` to compute the average for each variable. I used `kable()` from the `knitr` package to display the table and set `digits = 2` so the averages only show two decimal places. One issue I had to deal with was making sure the date column was converted with the correct format string, since the dates were written as month/day/year.

```{r}
#| include: false
library(tidyverse)
library(knitr)
```

```{r}
#| include: false
rivers <- read_csv("Labs/Data/rivers.csv", show_col_types = FALSE) %>%
  mutate(
    date = as.Date(date, format = "%m/%d/%Y"),
    river = as.factor(site),
    temp = as.numeric(temperature),
    do = as.numeric(dissolved_oxygen),
    ph = as.numeric(pH),
    month = as.numeric(format(date, "%m"))
  )
```

```{r}
#| results: asis
avg_table <- rivers %>%
  group_by(river) %>%
  summarize(
    avg_temp = mean(temp, na.rm = TRUE),
    avg_do = mean(do, na.rm = TRUE),
    avg_ph = mean(ph, na.rm = TRUE)
  )

knitr::kable(
  avg_table,
  digits = 2,
  col.names = c("River", "Average Temperature", "Average Dissolved Oxygen", "Average pH")
)
```

```{r}
plot1_data <- rivers %>%
  filter(temp > 25) %>%
  count(river, name = "count")

ggplot(plot1_data, aes(x = river, y = count, fill = river)) +
  geom_col() +
  labs(
    title = "Number of Temperature Observations Above 25°F",
    x = "River",
    y = "Number of Observations"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| include: false
wqs <- read_csv("Labs/Data/wqs.csv", show_col_types = FALSE) %>%
  mutate(
    river = as.factor(site),
    month = as.numeric(month),
    standard = as.numeric(standard)
  )
```

```{r}
combined <- rivers %>%
  left_join(wqs, by = c("river", "month"))

plot2_data <- combined %>%
  filter(temp > standard) %>%
  count(river, name = "count")

ggplot(plot2_data, aes(x = river, y = count, fill = river)) +
  geom_col() +
  labs(
    title = "Number of Temperature Exceedances of Water Quality Standards",
    x = "River",
    y = "Number of Exceedances"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

#Code

```{r}
#| eval: false

library(tidyverse)
library(knitr)

rivers <- read_csv("Labs/Data/rivers.csv", show_col_types = FALSE) %>%
  mutate(
    date = as.Date(date, format = "%m/%d/%Y"),
    river = as.factor(site),
    temp = as.numeric(temperature),
    do = as.numeric(dissolved_oxygen),
    ph = as.numeric(pH),
    month = as.numeric(format(date, "%m"))
  )

wqs <- read_csv("Labs/Data/wqs.csv", show_col_types = FALSE) %>%
  mutate(
    river = as.factor(site),
    month = as.numeric(month),
    standard = as.numeric(standard)
  )

avg_table <- rivers %>%
  group_by(river) %>%
  summarize(
    avg_temp = mean(temp, na.rm = TRUE),
    avg_do = mean(do, na.rm = TRUE),
    avg_ph = mean(ph, na.rm = TRUE)
  )

knitr::kable(
  avg_table,
  digits = 2,
  col.names = c("River", "Average Temperature", "Average Dissolved Oxygen", "Average pH")
)

plot1_data <- rivers %>%
  filter(temp > 25) %>%
  count(river, name = "count")

ggplot(plot1_data, aes(x = river, y = count, fill = river)) +
  geom_col() +
  labs(
    title = "Number of Temperature Observations Above 25°F",
    x = "River",
    y = "Number of Observations"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

combined <- rivers %>%
  left_join(wqs, by = c("river", "month"))

plot2_data <- combined %>%
  filter(temp > standard) %>%
  count(river, name = "count")

ggplot(plot2_data, aes(x = river, y = count, fill = river)) +
  geom_col() +
  labs(
    title = "Number of Temperature Exceedances of Water Quality Standards",
    x = "River",
    y = "Number of Exceedances"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


