---
title: "Exam 1 Preparation"
author: "Eli Crowder"
date: Feb 10 2026
format: pdf
---

```{r}
#| include: false
library(tidyverse)
```

## Group 1 - Packages (base R):

Use `install.packages()` once when first installing a package and then run `library()` at the beginning of each fresh instance of code (new document, file, etc) to be able to use that package.

```{r}
#| eval: false
install.packages("_")
library(_)
```

## Group 2 - Plotting (ggplot2):

-   `aes` or "aesthetic" is used to define the mappings between variables and their visual properties

-   `ggplot` initializes the plot. `data` and `mapping` are both included inside this function.

-   `geom_bar`, `geom_histogram`, `geom_line`, and `geom_point` are all different types of plots the function can create (bar graph, histogram, line, and points) and follows the + after closing a `ggplot` function

-   `labs` follows a + sign after the geom and allows you to create titles, subtitles, axis labels, and captions among other things

### For example,

```{r}
#| eval: false
ggplot(
  data = _,
  mapping = aes(x = _, y = _)
  ) + 
  geom_bar() +
  labs(title = "_", x = "_", y = "_", caption = "_")
```

## Group 3 - Data manipulation (dplyr):

-   `arrange` sorts rows of a dataset by one or more variables

    -   `arrange(df, var)`

-   `count` counts the number of observations by group

    -   `count(var)`

-   `desc` is used inside of `arrange` to sort in descending order

    -   `arrange(desc(var))`

-   `distinct` keeps unique rows based on one or more variables

    -   `distinct(var)`

-   `filter` keeps rows that satisfy specific conditions

    -   `filter(var > 10)`

-   `group_by` groups the data so subsequent operations are done within groups

    -   `group_by(var)`

-   `mutate` adds or modifies columns without changing the number of rows

    -   `mutate(new_var = expression)`

-   `relocate` allows you to move columns to a new position

    -   `relocate(var, .before = other_var)`

-   `rename` allows you to rename columns using new_name = old_name

    -   `rename(new = old)`

-   `select` allows you to choose a subset of columns from a dataset

    -   `select(col1, col2)`

-   `summarize` computes summary statistics and reduces data to one row per group

    -   `summarize(mean_x = mean(x))`

## Group 4 - Tidying Data (tidyr):

-   `pivot_longer` turns multiple columns into rows [making data longer]{.underline}

    -   `pivot_longer(cols, names_to, values_to)`

-   `pivot_wider` turns rows into columns [making data wider]{.underline}

    -   `pivot_wider(names_from, values_from)`
    
### For example,

```{r}
# Original Wide Data
scores <- tibble(
  student = c("Alex", "Blair", "Casey"),
  exam1 = c(85, 90, 88),
  exam2 = c(87, 92, 91)
)
```

```{r}
scores_long <- scores |>
  pivot_longer(
    cols = starts_with("exam"),
    names_to = "exam",
    values_to = "score"
  )
```

```{r}
# Original Long Data
weather <- tibble(
  day = c("Mon", "Mon", "Tue", "Tue"),
  city = c("Bozeman", "Helena", "Bozeman", "Helena"),
  temp = c(70, 68, 72, 69)
)
```

```{r}
weather_wide <- weather |>
  pivot_wider(
    names_from = city,
    values_from = temp
  )
```


## Group 5 - Viewing Data (base R):

-   `c` combines values into a vector

-   `dim` returns the number of rows and columns for a dataset

-   `head` returns the first six rows of a dataset by default but can return n rows

-   `max` returns the largest value

-   `mean` returns the average value (numeric values)

-   `min` returns the smallest value

-   `names` returns the column (variable) names of the dataset

-   `ncol` returns the number of columns

-   `nrow` returns the number of rows

