---
title: "Homework 1"
author: "Eli Crowder"
format: pdf
---

```{r}
#| include: false
library(palmerpenguins)
library(ggplot2)
library(tidyverse)
```

## 1.4.3 Exercises

1.  Make a bar plot of `species` of `penguins`, where you assign `species` to the `y` aesthetic. How is this plot different? 

```{r}
ggplot(penguins, aes(y = species)) +
  geom_bar()
```

This plot is different because the bars are arranged on the y axis and extend horizontally instead of vertically.

2.  How are the following two plots different? Which aesthetic, `color` or `fill`, is more useful for changing the color of bars?

```{r}
#| eval: false
ggplot(penguins, aes(x = species)) +
  geom_bar(color = "red")

ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "red")
```

The two plots are different in the way that they assign `red`. In the first where `color = "red"`, the bars are outlined in red and in the second where `fill = "red"` the whole bar is red. Because of this, I think `fill` is more useful when changing the bar color. 

3.  What does the `bins` argument in `geom_histogram()` do?

The `bins` argument in `geom_histogram()` controls how many bars or "bins" the histogram uses to group the data. For example, if `bins = 30`, it divides the data range into 30 equal-width bins, then count observations in each bin.

4.  Make a histogram of the `carat` variable in the `diamonds` dataset that is available when you load the tidyverse package. Experiment with different binwidths. What binwidth reveals the most interesting patterns?

```{r}
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = .01)
```

The `binwidth` that I thought was most interesting was .01 because it included a lot more bars and I was able to see large spikes around very specific carats. I think this revealed more information rather than just a few bigger bins. 






