---
title: "Homework 2: Reference sheet"
author: "Eli Crowder"
date: Jan 27 2026
format: pdf
---

```{r}
#| include: false
library(ggplot2)
library(tidyverse)
```

# Basics

### Installing and Running Packages

```{r}
#| eval: false
install.packages() # to install it on your computer; do this step just once
library() # to load tidyverse in a specific R session so you can use it
```

### Chunk Options

| Option           | Runs code | Shows code | Shows output |
|------------------|-----------|------------|--------------|
| *(default)*      | +         | +          | +            |
| `eval: false`    | -         | +          | -            |
| `echo: false`    | +         | -          | +            |
| `output: false`  | +         | +          | -            |
| `include: false` | +         | -          | -            |
| `display: true`  | +         | +          | +            |

Common uses:

Setup chunks → include: false

Show results, hide code → echo: false

Show example code only → eval: false

# Quarto Formatting

*italic* **bold** ~~strikeout~~ `code`

superscript^2^ subscript~2~

[underline]{.underline} [small caps]{.smallcaps}

"\## Headings"

"\# 1st Level Header"

"\## 2nd Level Header"

"\### 3rd Level Header"

## Lists

-   Bulleted list item 1

-   Item 2

    -   Item 2a

    -   Item 2b

1.  Numbered list item 1

2.  Item 2. The numbers are incremented automatically in the output.

## Links and images

<http://example.com>

[linked phrase](http://example.com)

## Tables

| First Header | Second Header |
|--------------|---------------|
| Content Cell | Content Cell  |
| Content Cell | Content Cell  |

# Data Analysis

-   names() → see the variables in the dataset

-   head() → shows you the first 6 rows

-   nrow() → how many rows (observations) are there?

-   ncol() → how many columns (variables)?

-   dim() → nrow x ncol; dim = dimensions

-   ?function → opens help info

# ggplot (Graphs)

```{r}
#| eval: false
ggplot(
data = _,
mapping = aes(x = _, y = _)
) +
geom_point(aes(color = _)) + geom_smooth() + labs(caption = "_")
```

-   ggplot(...) → initializes the plot data → specifies data frame being used
-   mapping = aes(x = *, y =* )
    -   Defines the global aesthetics
    -   Maps variables to the x- and y-axes
    -   These mappings apply to all geoms, unless overridden later
-   \+ → adds layers
    -   The + operator adds layers or components to the plot
    -   Each layer builds on the previous one
-   geom_point() → scatterplot
-   geom_bar() → bargraph
-   aes(color = ) → color by variable (this layer only)
-   geom_smooth() → trend line labs() → titles, labels, caption

# dplyr (Data Manipulation)

-   df |> function()
  -   Reads as: “take df then do this”
  -   Avoids nested function calls
-   select() → choose columns
-   filter() → choose rows
-   arrange() → sort rows
-   count() → count observations
-   rename() → rename columns
-   relocate() → move columns

# In-class notes for 1/27

Scripts vs Quarto docs

-   scripts (.R file extension) do not have any text formatting, just code and code comments 


