---
title: "STAT 408 Lab 1: Data Viz with `ggplot`"
author: "Eli Crowder"
date: today
date-format: "01, 15, 2026"
format: pdf
embed-resources: true
---

1. Load the three packages you are using for this lab on the `penguins` dataset.

```{r}
library(ggplot2)
library(palmerpenguins)
library(ggthemes)
```



2. How many rows (observations) are in `penguins`? How many columns (variables)? What are the variables (column names) in this dataset? Answer these three questions using code covered in **lab01_notes.qmd**.

```{r}
nrow(penguins)
ncol(penguins)
names(penguins)
```

There are 344 rows, 8 columns, and the variable names are species, island, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g, sex, and year. 



3. What does the `bill_depth_mm` variable in the `penguins` data frame describe? Read the help for ?penguins to find out.

This variable holds a number denoting bill depth (millimeters).




4. Make a scatterplot of `bill_depth_mm` vs. `bill_length_mm.` That is, make a scatterplot with `bill_depth_mm` on the y-axis and `bill_length_mm` on the x-axis. Describe (in text, not code or code comments) the relationship between these two variables.

```{r}
ggplot(
data = penguins,
mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) +
geom_point()
```

There seems to be two clusters of data, a cluster of higher bill depth associated with lower bill length and a cluster of low bill depth associated with med/higher bill length. There also seems to be slight negative relationship. 



5. What happens if you make a scatterplot of `species` vs. `bill_depth_mm`? (Make and show this scatterplot, then discuss in text what happens.) What might be a better choice of geom for this plot? (See other options mentioned in Section 1.2.3.)

```{r}
ggplot(
data = penguins,
mapping = aes(x = species, y = bill_depth_mm)
) +
geom_point()
```

There are three "columns" of data that aren't vry pleasing to look at. A better choice for geom might be geom_boxplot because it would display the data better.



6. Why does the following code give an error? (Write your answer in text.) Fix it. (NOTE: you need to delete the line that says `#| eval: false` so that this code will run when you render your Quarto document.)

There are no assigned variables or an x or y axis to plot on. 

```{r}
ggplot(
data = penguins,
mapping = aes(x = island, y = bill_depth_mm)
) +
geom_point()
```



7. What does the `na.rm` argument do in `geom_point()`? What is the default value of the argument? (Hint: to answer these questions, look up the help page for `geom_point()` the same way you did for the `penguins` dataset, and don't include the parentheses when you do that.) Then create a scatterplot where you successfully use this argument set to `TRUE`.

If FALSE, the default, missing values are removed with a warning. If TRUE, missing values are silently removed.


```{r}
ggplot(
data = penguins,
mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) +
geom_point(na.rm = TRUE)
```



8. Recreate the visualization in Exercise 8 of Section 1.2.5. What aesthetic should `bill_depth_mm` be mapped to? And should it be mapped at the global level or at the geom level?

We should have bill_depth_mm mapped to the color aesthetic. After trying, it can be mapped at either the global level or the geom level. 

```{r}
ggplot(
data = penguins,
mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
geom_point(aes(color = bill_depth_mm)) + geom_smooth()
```



9. Add the following caption to the plot you made in the previous exercise: "Data come from the palmerpenguins package." Hint: Take a look at the documentation for` labs()`.

```{r}
ggplot(
data = penguins,
mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
geom_point(aes(color = bill_depth_mm)) + geom_smooth() + labs(caption = "Data come from the palmerpenguins package.")
```



10. Will these two graphs look different? Why/why not? (Please leave the `#| eval: false` option in this code chunk.)

```{r}
#| eval: false

ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )
```

These two graphs will not look different, one sets everything globally and the other sets everything in the geom, just using the same data set and mapping each time. 



11. Type the name of the function we used in this lab for each of the following tasks.

- Getting the number of variables in a dataset: ncol()
- Setting up plot aesthetics: aes()
- Loading a package: library()
- Getting the number of rows in a dataset: nrow()
- Installing a package: install.packages()



12. Make a few notes here for yourself about some useful tips and tricks you learned (or were reminded of) in today's lab.

I have learned everything I know about R from this lab! Something that sticks out is global vs geom. Learning about how to make, color, and label graphs was also very interesting. 




