---
title: "STAT 408 Lab 4: Putting it all together"
author: "Eli Crowder"
format: pdf
---

```{r}
#| include: false
library(tidyverse)
```

### Question 1

```{r}
#| output: false
fish_encounters <- read_csv(
  "fish_encounters.csv",
  skip = 3
  
  ) |>
  rename(
    station = `Station name`,
    fish_id = `fish ID`
  )

```

In order to properly read in the data, I had to skip three lines because they included a discription, the author of the data, and then the source of the data. I also renamed the headers so that they would be easier to work with in the future.

### Question 2

```{r}
#| include: false
nrow(fish_encounters)
ncol(fish_encounters)
names(fish_encounters)
colSums(is.na(fish_encounters))
unique(fish_encounters$seen)
```

In the `fish_encounters` data set there are 114 rows, and 3 columns with `fish_id`, `station`, and `seen` as column names. There are zero values in any of the columns containing "N/A", and the only value recorded in the `seen` column is 1 since the observations were only recorded for fish that were seen.

### Question 3

```{r}
total_seen <- sum(fish_encounters$seen, na.rm = TRUE)
total_stations <- n_distinct(fish_encounters$station)

avg_fish_per_station <- total_seen/total_stations

station_summary <- fish_encounters |>
  group_by(station) |>
  summarise(
    total_seen = sum(seen, na.rm = TRUE)
  )

total_stations
avg_fish_per_station
station_summary
```

I computed how many different stations are recording fish with the average number of fish that were recorded at a station. I also recording the number of seen fish at each station.

### Question 4

```{r}
#| eval: false
ggplot(
  data = station_summary, 
  mapping = aes(x = station, y = total_seen)
  
) + 
  geom_col() +
  geom_hline(
    yintercept = avg_fish_per_station
  ) +
    labs(
      title = "Fish Seen per Station",
      x = "Station", 
      y = "Fish Seen"
    )
         
```

This graph combines the two metrics I calculated. It shows a bar graph with each station on the x axis and the amount of fish each station saw on the y axis. Then, it adds a horizontal line showing where the average is.

### Question 5

Overall, this data doesn't include much to work with. It would be really interesting if it included locations of the stations, types of fish, or anything like that. Something I could do however is see if a specific fish visited only specific stations. I'm not sure how I could do that yet.
