---
title: "STAT 408 Lab 2: dplyr"
author: "Eli Crowder"
date: Jan 22 2026
format: pdf
---

```{r}
#| include: false
library(nycflights13)
library(tidyverse)
```


## Section 3.2.5: working with rows

1. In a single pipeline for each condition, find all flights that meet the condition:

* Had an arrival delay of two or more hours
* Flew to Houston (`IAH` or `HOU`)
* Were operated by United, American, or Delta
* Departed in summer (July, August, and September)
* Arrived more than two hours late but didn’t leave late
* Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
#| output: false
flights |>
  filter(arr_delay >= 120)

flights |>
  filter(dest == "IAH" | dest == "HOU") 

flights |>
  filter(carrier == "UA" | carrier == "AA" | carrier == "DL")

flights |>
  filter(month == 7 | month == 8 | month == 9)

flights |>
  filter(arr_delay > 120, dep_delay <= 0)

flights |>
  filter(dep_delay >= 60, (dep_delay - arr_delay) > 30)
```


2. Sort `flights` to find the flights with the longest departure delays. Find the flights that left earliest in the morning.

```{r}
#| output: false
flights |>
  arrange(desc(dep_delay))

flights |>
  arrange(dep_time)
```


3. Sort `flights` to find the fastest flights. (Hint: Try including a math calculation inside of your function.)

```{r}
#| output: false
flights |>
  arrange(distance/air_time)
```


4. Was there a flight on every day of 2013?

```{r}
#| output: false
flights |>
  count(month, day)
```

Yes, there was a flight every day of 2013 because there are 365 rows. 

5. Which flights traveled the farthest distance? Which traveled the least distance?

```{r}
#| output: false
flights |>
  arrange(desc(distance))

flights |>
  arrange(distance)
```

Quite a few flights traveled 4,983 miles while one flight traveled 17 miles.

6. Does it matter what order you used `filter()` and `arrange()` if you’re using both? Why/why not? Think about the results and how much work the functions would have to do. 

I don't think the order matters because in most cases, the order of filter() and arrange() does not change the final result because filtering removes rows and arranging only reorders them.


## Section 3.3.5: working with columns

1. Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?

Well, `dep_time` is the time the plane actually takes off, `sched_dep_time` is the time it was supposed to take off, and `dep_delay` is the difference between the two where a positive means it took off later than schedules and a negative before it was scheduled. 

2. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from `flights`.

```{r}
#| output: false
flights |>
  select(dep_time, dep_delay, arr_time, arr_delay)

flights |>
  select(4, 6, 7, 9)

flights |>
  select(starts_with("dep"), starts_with("arr"))
```


3. What happens if you specify the name of the same variable multiple times in a `select()` call?

```{r}
#| output: false
flights |>
  select(dep_time, dep_time)
```

If you specify the name of the same variable multiple times it only returns the variable once. 

4. What does the `any_of()` function do? Why might it be helpful in conjunction with this vector?

```{r}
#| display: false
variables <- c("year", "month", "day", "dep_delay", "arr_delay")
```

The function `any_of()` returns columns if they exist but doesn't return an error if they don't. I don't really see how it could be helpful because all these columns exist in the data set. It could be helpful in the future if I am using this code again but I don't think that's the case. 

5. Does the result of running the following code surprise you? How do the select helpers deal with upper and lower case by default? How can you change that default?

```{r}
#| output: false
flights |> select(contains("TIME"))
```

No it doesn't surprise me it just returned all the columns that contain "time." The select helpers are case insensitive by default. You could use `ignore.case = FALSE` to ignore the default setting. 

6. Rename `air_time` to `air_time_min` to indicate units of measurement and move it to the beginning of the data frame.

```{r}
#| output: false
flights |>
  rename(air_time_min = air_time) |> 
  relocate(air_time_min)

```


7. Why doesn’t the following work, and what does the error mean?

```{r}
#| eval: false
flights |> 
  select(tailnum) |> 
  arrange(arr_delay)
#> Error in `arrange()`:
#> ℹ In argument: `..1 = arr_delay`.
#> Caused by error:
#> ! object 'arr_delay' not found
```

It doesn't work because everything is removed from the data frame but `tailnum` after the first function. The error is saying it can't "arrange" it because it's not there. 

