---
title: "Activity 2"
format: html
---

```{r}
#| include: false
library(palmerpenguins)
library(ggplot2)
```

## 1.3 ggplot2 calls

As we move on from these introductory sections, we’ll transition to a more concise expression of ggplot2 code. So far we’ve been very explicit, which is helpful when you are learning:

```{r}
#| eval: false
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```

Typically, the first one or two arguments to a function are so important that you should know them by heart. The first two arguments to ggplot() are data and mapping, in the remainder of the book, we won’t supply those names. That saves typing, and, by reducing the amount of extra text, makes it easier to see what’s different between plots. That’s a really important programming concern that we’ll come back to in \[Chapter 25\].

Rewriting the previous plot more concisely yields:

```{r}
#| eval: false
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()
```

In the future, you’ll also learn about the pipe, \|\>, which will allow you to create that plot with:

```{r}
#| eval: false
penguins |> 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()
```
